*DCVtestkit*: an R package for analysis of quality control dilution curves

### 2022-10-13 Manuscript/manuscript.qmd

Dilution curves have been used in many areas of biomedical and pharmaceutical sciences. In the field of metabolomics and lipidomics, quality control dilution curves for each molecule are plotted, as proposed by @croixmarie2009 to check if the sample matrix is the same among all samples. Molecules that do not have a linear relationship are rejected in spite of its relevance to the study. A cut off Pearson Correlation Coefficient value of $0.8$ was used to balance the risk of accepting spurious signals and removing signals related to important molecules. As such, many laboratories use commercial software to plot these dilution curves and report their corresponding off Pearson Correlation Coefficient value. These includes general purpose programs such as Excel and GraphPad Prism. Nevertheless, these general-purpose tools have their limitations.

While the Pearson Correlation Coefficient $R$ and/or Coefficient Of Determination $R^2$ are popular indicators of how well the dilution points fits a linear line, @sonnergaard2006 assets that these are not effective standalone numeric parameter to linearity. Using a modified version of Anscombe's quartet, the paper showed that a high coefficient of determination does not reflect a curve relationship well. Both @croixmarie2009 and @sands2021 support this view by indicating that an obvious non-linear response at the high and/or low ends of the diluted quality control sample dilution curve can still yield a high correlation coefficient. This is in agreement with @sanchez2021 findings of several saturated calibration curves with a $R$ value greater than $0.995$.

Fortunately, researcher who have studied deeply on model validation and misspecification have provided alternative graphical and statistical methods to improve the reliability of linearity assessment. For example, @vanloco2002 shared the presence of alternative statistics such as the Lack-of-fit test and quality coefficient that can be used to make better judgement on the shape of the dilution curve. While the progress in finding better metrics and algorithms to evaluate the linearity of dilution curve is commendable, these findings are rarely implemented in many general-purpose software.

In addition, with the advancement of technology in molecular identification coupled with a high emphasis on high throughput analysis and automation, today's targeted meatbolomics/lipidomics analysis can go as high as five hundred molecules being measured. This number can increase further in an untargeted approach. Using general-purpose programs to plot dilution curves for every single molecule may be tedious. Moreover, the report generated by these programs are static dilution plots coupled with statistical summary table. Having the analyst to look at five hundred dilution curves individually is time consuming and inefficient in understanding the data.

To address the issues mentioned above, we have created an R package, `DCVtestkit`, to assist analysts, not only to plot these dilution curves for many molecules easily, but also provided additional statistical results, other than the Pearson Correlation Coefficient $R$ or coefficient of determination $R^2$, for these curves. These alternative statistical results may be helpful in describing the shape the dilution curve possess. The R package also provide several ways of reporting the dilution data, including one as an interactive html file. This allows the analyst to group, filter and sort the dilution plots, to only look at the relevant ones, like saturated dilution curves, and identify problematic molecules quickly.

@@

Here, we introduce the workflow in which R package, `DCVtestkit`, follows.

>>>>

Dilution curves have been used in many areas of biomedical and pharmaceutical sciences. In the field of metabolomics and lipidomics, dilution curves for each molecule measured, usually in a quality control sample generated by pooling equal aliquots from each study sample, are plotted as proposed by @croixmarie2009 to check if the sample matrix is the same among all samples. Molecules that do not show a linear dilution curve can be rejected before further statistical analysis, in spite of their relevance to the study, as their quantitation in different sample groups might be compromised. A cut off Pearson Correlation Coefficient value of $0.8$ is commonly used to balance the risk of accepting or removing signals related to relevant molecules. As such, many laboratories use commercial software to plot these dilution curves and report their corresponding Pearson Correlation Coefficient value. These include either mass spectrometry vendors’ software or, more in general, programs such as Excel and GraphPad Prism. Nevertheless, these tools have limitations.

While the Pearson Correlation Coefficient $R$ and/or Coefficient Of Determination $R^2$ are popular indicators of how well the dilution points fits a linear line, @sonnergaard2006 estimated that these are not effective standalone numeric parameter to estimate linearity. Using a modified version of Anscombe’s quartet, the paper showed that a high coefficient of determination does not reflect well a curve relationship. Both @croixmarie2009 and @sands2021 support this view, by indicating that an obvious non-linear response at the high and/or low ends of the quality control sample-based dilution curve can still yield a high correlation coefficient. This is in agreement with @sanchez2021 findings of several saturated calibration curves with $R$ values greater than $0.995$.

Fortunately, researchers who have studied deeply model validation and misspecification have provided alternative graphical and statistical methods to improve the reliability of linearity assessments. For example, @vanloco2002 highlighted the presence of alternative statistics, such as the Lack-of-fit test and quality coefficient, that can be used to make better judgement on the shape of the dilution curve. While the progress in finding better metrics and algorithms to evaluate the linearity of dilution curves is commendable, these findings are rarely implemented in more general-purpose software.

In addition, with the advancement of the technology used for molecular identification coupled with a high emphasis on high throughput analysis and automation, today’s targeted metabolomics/lipidomics can include more than eight hundred analytes measured by a single method. This number can also increase further in an untargeted approach. In these cases, using general-purpose programs to plot dilution curves for every single molecule may be tedious. Moreover, the reports generated by these programs are static dilution plots coupled with a statistical summary table. Having the analyst to look at five hundred dilution curves individually is time consuming and inefficient for data evaluation.

To address the issues mentioned above, we have created an R package, DCVtestkit, to assist analysts, not only to plot these dilution curves for many molecules easily but also providing additional statistical results, other than the Pearson Correlation Coefficient $R$ or coefficient of determination $R^2$, for these curves. These alternative statistical results may be helpful in better describing the shape of the dilution curve. The R package also provides several ways of reporting the dilution data, including an interactive html file. This allows the analyst to group, filter and sort the dilution plots, to only look at the relevant ones, such as the saturated dilution curves, and identify problematic molecules quickly.

@@

Here, we introduce the workflow in which R package, DCVtestkit.
